---
title: "Simple RIME Report"
author: Ashley Sawle
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{SimpleRIMEReport}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteKeywords{Proteomics, TMT}
  \usepackage[utf8]{inputenc}
---

```{r setup, echo=FALSE, results="hide"}
knitr::opts_chunk$set(tidy = FALSE,
                      cache = FALSE,
                      dev = "png",
                      message = FALSE, error = FALSE, warning = TRUE)
``` 

This document provides brief tutorial of the _SimpleRIMEReport_ package.

The package generates a report in the form of an Excel workbook from the Protein
and Peptides output of Proteome Discoverer.

There are two steps to the process:

1) Generate a sample sheet
2) Generate the report

After generating the sample sheet it is important to check the details to make 
sure they are as you need them to be. In particular, the package will attempt
to identify the correct UniProt ID for bait proteins, however, in some cases 
the same Gene Symbol will be associated with multiple UniProt IDs and you should
select the correct one. Furthermore, if a custom database has been used, then 
the package will not able to identify the accession ID for the protein.

# 1. Generate sample table

The function `createSampleTable` requires the directory that contains the PD
outputs. It uses the file names to generate a sample table.

The sample table will be  saved in the data directory as
"XXXXX_sample_details.csv"

The sample table will contain 5 columns:  

* ProteinFile - the file name for 'protiengroups' data  
* PeptideFile - the file name for 'psms' data  
* SampleName - the name for the sample - this is used for the plot titles and
             the names of the worksheets  
* BaitProteinName - the protein name for the bait protein, e.g. ER, FOXA1 etc.
                  This is only used in the plot titles, it could be anything  
* BaitProteinUniProtID - the UniProt ID for the bait protein e.g. P03372, P55317
                      this is used to look up the fasta sequence  
                      
The function `createSampleTable` will attempt to guess the SampleName and
BaitProteinName from the the protein file name.

The assumed structure of the protein file name is:  

QE\_HF\_\<Date\>\_\<PR_number\>\_\<Initials\>\_\<SampleName\>\_\<BaitProteinName\>\_proteingroups.txt

The SampleName can include "_" e.g.:  

  QE_HF_04092018_PR1025_SAN_Sample1_VEH_ARID2_proteingroups.txt

gives:

  SampleName: "Sample1_VEH"  
  BaitProteinName: "ARID2"  

You'll need to check the sample sheet and modify as necessary before
proceeding. The sample table can either be modified directly in R, or you can
open the saved version in a spreadsheet (e.g. Excel) and edit it and the load it
back into R.

```{r eval=FALSE}
library(SimpleRIMEReport)
sampleData <- createSampleTable("My_Data_Directory")
```

This will create a table called `sampleData` and will save a copy of this to 
`My_Data_Directory/My_Data_Directory_sample_details.csv`.

If you edit the report in Excel, you will then need to reload the sample table:

```{r eval=FALSE}
library(readr)
sampleData <- read_csv("My_Data_Directory/My_Data_Directory_sample_details.csv")
```

# 3. Generate the report

Once you have adjusted the sample table you can use the function `createReport`
to generate the report.

```{r eval=FALSE}
createReport(sampleData)
```

This will create the report called: 
`My_Data_Directory/My_Data_Directory.Simple_RIME_report.xlsx`.

You can optionally provide an alternative name for the report e.g.:

```{r eval=FALSE}
createReport(sampleData, outputFileName = "RIME_reports/A_New_RIME_Report.xlsx")
```

This would create the report called `A_New_RIME_Report.xlsx` in the directory 
`RIME_reports`. Note that the directory "RIME_reports" would not be created
by the function, it would have to already exist.

## Session Information

```{r info,echo=TRUE}
sessionInfo()
```

