---
title: "Simple RIME Report"
author: Ashley Sawle
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{SimpleRIMEReport}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteKeywords{Proteomics, TMT}
  \usepackage[utf8]{inputenc}
---

```{r setup, echo = FALSE, results = "hide"}
knitr::opts_chunk$set(tidy = FALSE,
                      cache = FALSE,
                      dev = "png",
                      message = FALSE, error = FALSE, warning = TRUE)
```

This document provides brief tutorial of the _SimpleRIMEReport_ package.

The package generates a report in the form of an Excel workbook from the Protein
and Peptides output of Proteome Discoverer.

There are two steps to the process:

1) Generate a sample sheet
2) Generate the report

## 1. Generate sample table

The sample table needs to contain 6 columns:

* ProteinFile - the file name for 'protein groups' data
* PeptideFile - the file name for 'psms' data
* SampleName - the name for the sample - this is used for the plot titles and
             the names of the worksheets
* BaitProteinName - the protein name for the bait protein, e.g. ER, FOXA1 etc.
                  This is only used in the plot titles, it could be anything,
                  however, if there is no bait protein it should be one of
                  "IgG", "Empty", "None", or "Control".
* BaitProteinUniProtID - the UniProt ID for the bait protein e.g. P03372, P55317
                      this is used to look up the fasta sequence. If there is
                      no bait protein, then this should be left empty.
* SampleType - whether the sample is a "control" sample or something else.
              This is used in determining which samples are used to identify
              background proteins. 

The sample table can be created by hand using a spreadsheet, alternatively, the
package provides the function `createSampleTable` to help you generate a
suitable sample table.

The function `createSampleTable` requires the directory that contains the PD
outputs. It uses the file names to generate a sample table.

The function `createSampleTable` will attempt to guess the SampleName and
BaitProteinName from the the protein file name. The function recognises outputs
from Protein Discoverer version 1.4 and version 2.4.

For PD version 1.4 filenames are expected to have the following naming convention:

`QE_HF_<Date>_<PR_number>_<Initials>_<SampleName>_<BaitProteinName>_proteingroups.txt`
`QE_HF_<Date>_<PR_number>_<Initials>_<SampleName>_<BaitProteinName>_psms.txt`

For PD version 2.4 filenames are expected to have the following naming convention:

`QE_HF_<Date>_<PR_number>_<Initials>_<SampleName>_<BaitProteinName>_Proteins.txt`
`QE_HF_<Date>_<PR_number>_<Initials>_<SampleName>_<BaitProteinName>_PeptideGroups.txt`

The SampleName can include "_" e.g.:

`QE_HF_04092018_PR1025_SAN_Sample1_VEH_ARID2_proteingroups.txt`

gives:

* SampleName: "Sample1_VEH"  
* BaitProteinName: "ARID2"

After generating the sample sheet it is important to check the details to make
sure they are as you need them to be. In particular, the package will attempt
to identify the correct UniProt ID for bait proteins, however, in some cases
the same Gene Symbol will be associated with multiple UniProt IDs and you should
select the correct one. Furthermore, if a custom database has been used, then
the package will not able to identify the accession ID for the protein.

The sample table will be  saved in the data directory as
"XXXXX_sample_details.csv" where XXXXX is the name of the data directory.

If the file names do not follow the above convention, then you'll need to modify
the sample sheet as necessary before proceeding. The sample table can either be
modified directly in R, or you can open the saved version in a spreadsheet (e.g.
Excel) and edit it and the load it back into R.

```{r eval = FALSE}
library(SimpleRIMEReport)
sampleData <- createSampleTable("My_Data_Directory")
```

This will create a table called `sampleData` and will save a copy of this to
`My_Data_Directory/My_Data_Directory_sample_details.csv`.

If you edit the report in Excel, you will then need to reload the sample table:

```{r eval = FALSE}
library(readr)
sampleData <- read_csv("My_Data_Directory/My_Data_Directory_sample_details.csv")
```

## 2. Generate the report

Once you have adjusted the sample table you can use the function `createReport`
to generate the report.

If the "Master" column is present, the proteins are filtered to only keep
proteins marked "IsMasterProtein".

```{r eval = FALSE}
createReport(sampleData)
```

This will create the report called:
`My_Data_Directory/My_Data_Directory.Simple_RIME_report.xlsx`.

You can optionally provide an alternative name for the report e.g.:

```{r eval = FALSE}
createReport(sampleData, outputFileName = "RIME_reports/A_New_RIME_Report.xlsx")
```

This would create the report called `A_New_RIME_Report.xlsx` in the directory
`RIME_reports`. Note that the directory "RIME_reports" would not be created
by the function, it would have to already exist.

### Annotation

The packages uses an internal copy of the SwissProt FASTA file to annotate the
proteins in the report. This is stored in the package and was last update 15th
July 2025.

### Custom FASTA files

If custom FASTA files have been used in the Protein Discoverer analysis, these
should also be provided to the `createReport` function so that the proteins can
be correctly annotated and so that the amino acid sequence is available for the
plot.

```{r eval = FALSE}
createReport(sampleData, additionalFasta = "My_Data_Directory/My_Protein.fasta")
```

The Accession, Gene ID and description for the annotation will be extracted
from the header for each sequence; the expected format is:

`>{Database}|{Accession}|{GeneID} {Description}`

This is the standard format of the header in FASTA files downloaded from
UniProt, for example:

`>sp|NUP98-HOXA9|NUP98-HOXA9_HUMAN OS = Homo sapiens`

where `sp` means this is a SwissProt entry, the Accession is `NUP98-HOXA9` and
the Gene ID is `NUP98-HOXA9_HUMAN`. Any further information is add to the
Description.

Accessions **must** be unique within the annotation. If the additional annotation
derived from the FASTA file contains accessions that are already present in SwissProt
these will be dropped from the internal SwissProt annotation on the basis that the
(presumably) modified entry in the additional FASTA is required.

The new annotation will be checked for duplicated or missing accessions.

## Session Information

```{r info,echo = TRUE}
sessionInfo()
```

