---
title: "Simple RIME Report"
author: Ashley Sawle
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{SimpleRIMEReport}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteKeywords{Proteomics, TMT}
  \usepackage[utf8]{inputenc}
---

```{r setup, echo = FALSE, results = "hide"}
knitr::opts_chunk$set(tidy = FALSE,
                      cache = FALSE,
                      dev = "png",
                      message = FALSE, error = FALSE, warning = TRUE)
```

This document provides brief tutorial of the _SimpleRIMEReport_ package.

The package generates a report in the form of an Excel workbook from the Protein
and Peptides output of Proteome Discoverer.

There are two steps to the process:

1) Generate a sample table
2) Generate the report

## 1. Generate sample table

The main input for the report is a sample table that contains the details of
the samples that you want to include in the report. The sample table is a
data frame that contains the details of each sample.

The sample table can be created by hand or generated using the function
`createSampleTable`.

### Sample table format

The sample table needs to contain 6 columns:

* **ProteinFile** - the file name for 'protein groups' data
* **PeptideFile** - the file name for 'psms' data
* **SampleName** - the name for the sample - this is used for the plot titles and
                   the names of the worksheets
* **BaitProteinName** - the protein name for the bait protein, e.g. ER, FOXA1 etc.
                        This is only used in the plot titles, it could be
                        anything, however, if there is no bait protein it should
                        be one of "IgG", "Empty", "None", or "Control".
* **BaitProteinUniProtID** - the UniProt ID for the bait protein, e.g. P03372,
                             P55317.  This is used to look up the fasta
                             sequence. If there is no bait protein, then this
                             should be left empty.
* **SampleType** - whether the sample is a "Control" sample or something else.
                   The samples marked as "Control" are used to identify
                   background proteins. 

### Creating the sample table

The sample table can be created by hand using a spreadsheet (e.g. Excel) and
then loaded into R. Alternatively, the package provides the function
`createSampleTable` to help you generate a suitable sample table. 

The `createSampleTable` function requires the directory that contains the PD
outputs.

The sample table will be  saved in the data directory as
"XXXXX_sample_details.csv" where XXXXX is the name of the data directory.

For example:

```{r eval = FALSE}
library(SimpleRIMEReport)
sampleData <- createSampleTable("My_Data_Directory")
```

This will create a data.frame called `sampleData` and will save a copy of this
to `My_Data_Directory/My_Data_Directory_sample_details.csv`.

The function uses the names of files in the directory to populate the sample
table.  The function will attempt to guess the SampleName and BaitProteinName
from the the protein file name. The function recognises outputs from Protein
Discoverer version 1.4 and version 2.4.

For PD version 1.4 filenames are expected to have the following naming
convention:

`QE_HF_<Date>_<PR_number>_<Initials>_<SampleName>_<BaitProteinName>_proteingroups.txt`
`QE_HF_<Date>_<PR_number>_<Initials>_<SampleName>_<BaitProteinName>_psms.txt`

For PD version 2.4 filenames are expected to have the following naming convention:

`QE_HF_<Date>_<PR_number>_<Initials>_<SampleName>_<BaitProteinName>_Proteins.txt`
`QE_HF_<Date>_<PR_number>_<Initials>_<SampleName>_<BaitProteinName>_PeptideGroups.txt`

The SampleName can include "_" e.g.:

`QE_HF_04092018_PR1025_SAN_Sample1_VEH_ARID2_proteingroups.txt`

gives:

* SampleName: "Sample1_VEH"  
* BaitProteinName: "ARID2"

### Check the contents of the sample table 

After generating the sample sheet it is important to check the details to make
sure they are as you need them to be.

In particular, the package will attempt to identify the correct UniProt ID for
bait proteins, however, in some cases the same Gene Symbol will be associated
with multiple UniProt IDs and you should select the correct one. Furthermore, if
a custom database has been used, then the package will not able to identify the
accession ID for the protein.

In addition, the package will do it's best to fill in the SampleType
column, however, it may not always be able to do this correctly. The SampleType
column is used to identify the control samples, so it is important that this
column is correct.

If the file names do not follow the above convention, then you'll definitely
need to modify the sample sheet as necessary before proceeding. The sample table
can either be modified directly in R, or you can open the saved version in a
spreadsheet (e.g.  Excel), edit it and then load it back into R.

If you edit the sample table in Excel, you will then need to reload it:

```{r eval = FALSE}
library(tidyverse)
sampleData <- read_csv("My_Data_Directory/My_Data_Directory_sample_details.csv")
```

## 2. Generate the report

Once you have adjusted the sample table you can use the function `createReport`
to generate the report.

```{r eval = FALSE}
createReport(sampleData, dataDir = "My_Data_Directory")
```

This will create the report called:
`My_Data_Directory/My_Data_Directory.Simple_RIME_report.xlsx`.

You can optionally provide an alternative name for the report e.g.:

```{r eval = FALSE}
createReport(sampleData,
             dataDir = "My_Data_Directory",
             outputFileName = "RIME_reports/A_New_RIME_Report.xlsx")
```

This would create a report called `A_New_RIME_Report.xlsx` in the directory
`RIME_reports`. Note that the directory "RIME_reports" would not be created
by the function, it would have to already exist.

### Annotation

The package includes an internal copy of the SwissProt FASTA file, which is used
to annotate the proteins in the report and to find sequence information for the
coverage plots. This is stored in the package and was last update 15th July
2025.

### Custom FASTA files

If custom FASTA files have been used in the Protein Discoverer analysis, these
should also be provided to the `createReport` function so that the proteins can
be correctly annotated and so that the amino acid sequence is available for the
plot.

```{r eval = FALSE}
createReport(sampleData,
             dataDir = "My_Data_Directory",
             additionalFasta = "My_Data_Directory/My_Protein.fasta")
```

The Accession, Gene ID and description for the annotation will be extracted
from the header for each sequence; the expected format is:

`>{Database}|{Accession}|{GeneID} {Description}`

This is the standard format of the header in FASTA files downloaded from
UniProt, for example:

`>sp|P55317|FOXA1_HUMAN Hepatocyte nuclear factor 3-alpha OS=Homo sapiens OX=9606 GN=FOXA1`

where `sp` means this is a SwissProt entry, the Accession is `P55317` and
the Gene ID is `FOXA1_HUMAN`. Any further information on the line is added to
the Description. The `{Database}` is not used by the function and could be
anything.

Accessions **must** be unique within the annotation. If the additional annotation
derived from the FASTA file contains accessions that are already present in SwissProt
these will be dropped from the internal SwissProt annotation on the basis that the
(presumably) modified entry in the additional FASTA is required.

The new annotation will be checked for duplicated or missing accessions.

--------------------------------------------------------------------------------