---
title: "Simple RIME Report"
author: Ashley Sawle
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{SimpleRIMEReport}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteKeywords{Proteomics, TMT}
  \usepackage[utf8]{inputenc}
---

```{r setup, echo = FALSE, results = "hide"}
knitr::opts_chunk$set(tidy = FALSE,
                      cache = FALSE,
                      dev = "png",
                      message = FALSE, error = FALSE, warning = TRUE)
```

This document provides brief tutorial of the _SimpleRIMEReport_ package.

The package generates a report in the form of an Excel workbook from the Protein
and Peptides output of Proteome Discoverer.

There are two steps to the process:

1) Generate a sample sheet
2) Generate the report

After generating the sample sheet it is important to check the details to make
sure they are as you need them to be. In particular, the package will attempt
to identify the correct UniProt ID for bait proteins, however, in some cases
the same Gene Symbol will be associated with multiple UniProt IDs and you should
select the correct one. Furthermore, if a custom database has been used, then
the package will not able to identify the accession ID for the protein.

## 1. Generate sample table

The function `createSampleTable` requires the directory that contains the PD
outputs. It uses the file names to generate a sample table.

The sample table will be  saved in the data directory as
"XXXXX_sample_details.csv"

The sample table will contain 5 columns:

* ProteinFile - the file name for 'protein groups' data
* PeptideFile - the file name for 'psms' data
* SampleName - the name for the sample - this is used for the plot titles and
             the names of the worksheets
* BaitProteinName - the protein name for the bait protein, e.g. ER, FOXA1 etc.
                  This is only used in the plot titles, it could be anything
* BaitProteinUniProtID - the UniProt ID for the bait protein e.g. P03372, P55317
                      this is used to look up the fasta sequence

The function `createSampleTable` will attempt to guess the SampleName and
BaitProteinName from the the protein file name. The function recognises outputs
from Protein Discoverer version 1.4 and version 2.4.

For PD version 1.4 filenames are expected to have the following naming convention:

`QE_HF_<Date>_<PR_number>_<Initials>_<SampleName>_<BaitProteinName>_proteingroups.txt`
`QE_HF_<Date>_<PR_number>_<Initials>_<SampleName>_<BaitProteinName>_psms.txt`

For PD version 2.4 filenames are expected to have the following naming convention:

`QE_HF_<Date>_<PR_number>_<Initials>_<SampleName>_<BaitProteinName>_Proteins.txt`
`QE_HF_<Date>_<PR_number>_<Initials>_<SampleName>_<BaitProteinName>_PeptideGroups.txt`

The SampleName can include "_" e.g.:

`QE_HF_04092018_PR1025_SAN_Sample1_VEH_ARID2_proteingroups.txt`

gives:

  SampleName: "Sample1_VEH"
  BaitProteinName: "ARID2"

You'll need to check the sample table and modify as necessary before
proceeding. The sample table can either be modified directly in R, or you can
open the saved version in a spreadsheet (e.g. Excel) and edit it and the load it
back into R.

```{r eval = FALSE}
library(SimpleRIMEReport)
sampleData <- createSampleTable("My_Data_Directory")
```

This will create a table called `sampleData` and will save a copy of this to
`My_Data_Directory/My_Data_Directory_sample_details.csv`.

If you edit the report in Excel, you will then need to reload the sample table:

```{r eval = FALSE}
library(readr)
sampleData <- read_csv("My_Data_Directory/My_Data_Directory_sample_details.csv")
```

## 2. Generate the report

Once you have adjusted the sample table you can use the function `createReport`
to generate the report.

If the "Master" column is present, the proteins are filtered to only keep
proteins marked "IsMasterProtein".

```{r eval = FALSE}
createReport(sampleData)
```

This will create the report called:
`My_Data_Directory/My_Data_Directory.Simple_RIME_report.xlsx`.

You can optionally provide an alternative name for the report e.g.:

```{r eval = FALSE}
createReport(sampleData, outputFileName = "RIME_reports/A_New_RIME_Report.xlsx")
```

This would create the report called `A_New_RIME_Report.xlsx` in the directory
`RIME_reports`. Note that the directory "RIME_reports" would not be created
by the function, it would have to already exist.

### Custom FASTA files

If custom FASTA files (non-UniProt) have been used in the Protein Discoverer
analysis, these should also be provided to the `createReport` function so that
the proteins can be correctly annotated and so that the amino acid sequence is
available for the plot.

```{r eval = FALSE}
createReport(sampleData, additionalFasta = "My_Data_Directory/My_Protein.fasta")
```

You can provide more than one fasta:


```{r eval = FALSE}
myFastaFiles <- c("My_Data_Directory/My_Protein1.fasta",
                  "My_Data_Directory/My_Protein2.fasta",
                  "My_Data_Directory/My_Protein3.fasta")
createReport(sampleData, additionalFasta = myFastaFiles)
```

The Accession, Gene ID and description for the annotation will be extracted
from the header for each sequence; the expected format is:

`>sp|<Accession>|<GeneID> <Description>`

e.g.:

`>sp|NUP98-HOXA9|NUP98-HOXA9_HUMAN OS = Homo sapiens`

where the Accession is `NUP98-HOXA9` and the Gene ID is `NUP98-HOXA9_HUMAN`. Any
further information is add to the Description.

## Session Information

```{r info,echo = TRUE}
sessionInfo()
```

